Regression Cube Analysis of Cohort Study Data

### 2015-03-25 draft.tex
added Ulis data description and revisions

Hielscher et al. \cite{Hielscher} assess subject similarity towards hepatic steatosis (fatty liver disease) to identify common risk factors using [...]?
\com{Uli, can you please complete this, I do not have the pdf available!}
%% Learning and inspecting classification rules from longitudinal epidemiological data to identify predictive features on hepatic steatosis
Niemann et al. \cite{Niemann2014} investigate risk factors of the hepatic steatosis using decision trees with a interactive data mining tools.
%%
They extracted association rules, which serve as basis for this works proof-of-concept evaluation.
%%
We try to reproduce their results and investigate their findings further.
%%
%\com{Glaßer, S., Niemann, U., Preim, B., and Spiliopoulou, M. (2013). Can we Distinguish Between Benign and Ma- lignant Breast Tumors in DCE-MRI}

@@

\subsection{Target-Variable-dependent Dimension Reduction}
\com{Uli, Can you please proofread this?}
%%
The vast majority of features in an epidemiological data have no or a very low relation towards the target feature.
%%
Identifying these features und excluding them from the calculation can reduce the number of dimension significantly.
%%
The \emph{Correlation based Feature Selection} (CFS) algorithm is very popular in data mining for this task \cite{CFS}.
%%
It uses information entropy to select the features which have the most explanatory power towards the target feature.
%%
At the same time it tries to reduce the correlation between the independent features.
%%
When for example the \emph{body weight} has a strong explanatory power towards the target, it is likely that \emph{BMI} or \emph{waist circumference} behave similar towards it.
%%
They, however, correlate strongly with each other.
%%
The CFS algorithm would then select the feature which has the largest explanatory power and discards the other dimensions.

@@

The data of each cohort was presented as individual \texttt{SPSS} files.
%%
All data related to the mammography was stored in a additional file.
%%
We converted the \texttt{SPSS} data sets to CSV and used \texttt{R} to merge the data sets together using their ID.
%%
All features were renamed to be expressive, e.g., \emph{chro\_09a} is now denoted as \emph{Disease\_Osteoporosis}.
%%
This avoids the need of defining a separate data dictionary file for the \emph{Regression Cube}, translating the feature names.
%%
All male subjects were removed as their data does not contribute to the analysis.
%%
\subsection{The Hepatic Steatosis Data Set}
\com{Uli: Describe the data set and also prior work on it.}
%%
\subsubsection{Data Preprocessing}
%%
\subsection{Case 1: Hypothesis-free Analysis of the Breast Cancer Data Set}
%%
\subsection{Case 2: Hypothesis-driven Analysis of the Hepatic Steatosis Data Set}
%%
\subsection{Further Feedback and Lessons Learned} \label{Lessons Learned}
%%
\com{Feedback from the evaluation goes here.}
\com{Time-aspect critical, interactive analysis requires for fast response. The method needs to be speeded up.}

>>>>

%Hielscher et al. \cite{Hielscher} assess subject similarity towards hepatic steatosis (fatty liver disease) to identify common risk factors using [...]?
%\com{Uli, can you please complete this, I do not have the pdf available!}
%% Learning and inspecting classification rules from longitudinal epidemiological data to identify predictive features on hepatic steatosis
Niemann et al. \cite{Niemann2014} investigate risk factors of the hepatic steatosis using decision trees with a interactive data mining tools.
%%
They extracted association rules, which serve as basis for this works proof-of-concept evaluation.
%%
In \cite{Niemann2015}, Niemann et al. improve on classification performance by generating features (called \emph{Evolution Features}) that describe latent temporal information across the study waves.
%%
We try to reproduce their results and investigate their findings further.
%%
%\com{Glaßer, S., Niemann, U., Preim, B., and Spiliopoulou, M. (2013). Can we Distinguish Between Benign and Ma- lignant Breast Tumors in DCE-MRI}

@@

\subsection{Target-Variable-dependent Dimension Reduction}
% \com{Uli, Can you please proofread this?}
% %%
% The vast majority of features in an epidemiological data have no or a very low relation towards the target feature.
% %%
% Identifying these features und excluding them from the calculation can reduce the number of dimension significantly.
% %%
% The \emph{Correlation based Feature Selection} (CFS) algorithm is very popular in data mining for this task \cite{CFS}.
% %%
% It uses information entropy to select the features which have the most explanatory power towards the target feature.
% %%
% At the same time it tries to reduce the correlation between the independent features.
% %%
In epidemiological studies, manifold recordings lead to an abundance of features.
%%
Generally many of them exhibit low or none correlation towards the target feature.
%%
Identifying irrelevant features and excluding them from the feature space considerably reduces computational costs and contributes to presentiveness of the Regression Cube.
%%
The Correlation-based feature selection (CFS) \cite{CFS} aims to find a feature subset that maximizes the \emph{merit value} $M_F$ which is the ratio between the average feature-class and feature-feature dependencies of features in F.
%%
The dependency of a set of features utilizes the entropy-based information gain to measure the explanatory power towards the target feature.
%%
Beginning with an empty set of features $F$ the CFS algorithm iteratively adds the feature $f$ to $F$ that leads to the highest new merit-value $M_{F\cup f}$ and halts when adding any feature does not increase the merit.
%%
When for example the \emph{body weight} has a strong explanatory power towards the target, it is likely that \emph{BMI} or \emph{waist circumference} behave similar towards it.
%%
They, however, correlate strongly with each other.
%%
The CFS algorithm would then select the feature which has the largest explanatory power and discards the other dimensions.

@@

The data of each cohort was presented as individual \texttt{SPSS} files.
%%
All data related to the mammography was stored in a additional file.
%%
We converted the \texttt{SPSS} data sets to CSV and used \texttt{R} to merge the data sets together using their ID.
%%
All features were renamed to be expressive, e.g., \emph{chro\_09a} is now denoted as \emph{Disease\_Osteoporosis}.
%%
This avoids the need of defining a separate data dictionary file for the \emph{Regression Cube}, translating the feature names.
%%
All male subjects were removed as their data does not contribute to the analysis.
%%
\subsection{The Hepatic Steatosis Data Set}
We employ a subset of the data set which was used in \cite{Niemann2014} to identify predictive features with respect to the reversible disorder hepatic steatosis (fatty liver).
%%
The binary target feature is derived from the liver fat concentration which was measured with Magnetic Resonance Imaging (MRI).
%%
In particular, participants with a liver fat concentration of no more than 10\% are mapped to the 'negative' class; values greater than 10\% are mapped to the 'positive' class.
%%
The data set contains labels for 578 participants for which the MRI data was available to that point.
%%
Note that MRI was performed only in SHIP-2, so the target feature is present for just one of three study waves.

%%
Apart from the target feature, the data set contains 199 features extracted from participants' questionnaire answers and medical tests.
%%
They are features on socio-demographics (sex, age, etc.), features on consumption behavior (e.g. alcohol and cigarettes), SNPs, features extracted from laboratory data (e.g. sera concentrations), and two features on the results of the liver ultrasound.
%%
The names of 85 features have the suffix \emph{s0} meaning that their values have been recorded in SHIP-0 (first study moment), 50 features exhibit the suffix \emph{s1} and 55 features have the suffix \emph{s2}, alongside 10 time-independent SNPs.
%%

In \cite{Niemann2014}, the authors show that women and men exhibit differences in the class distribution of the liver fat concentration.
%%
Also, for women they identify that there is a association between age and liver fat concentration.
%%
They compute an appropriate cut-off value of 52 years for which the class distribution is most homogeneous within the resulting subsets.
%%
Based on this observations, we perform our analysis on three sub-populations, $PartitionF$, $PartitionM$ and \emph{$F:age\>52$}.
%%
\subsection{Case 1: Hypothesis-free Analysis of the Breast Cancer Data Set}
%%
\subsection{Case 2: Hypothesis-driven Analysis of the Hepatic Steatosis Data Set}
%%
\subsection{Further Feedback and Lessons Learned} \label{Lessons Learned}
%%
\com{Feedback from the evaluation goes here.}
\com{Time-aspect critical, interactive analysis requires for fast response. The method needs to be speeded up.}
